### 2016-01-11 doc/thesis/intro.tex
Reading and writing

\subsection{Overview of phylodynamics}

@@

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa}). Phylogenies consist of nodes,
representing taxa, and edges or branches, which connect taxa to their
evolutionary ancestors and descendants. The \defn{tips} of a phylogeny, that
is, the nodes without any descendants, correspond to \defn{extant}, or
observed, taxa, while the \defn{internal nodes} correspond to their extinct
common ancestors. Phylogenies may have a \defn{root}, which is a node with no
descendants distinguished as the most recent common ancestor of all the extant
taxa. When such a root exists, the tree is referred to as being \defn{rooted};
otherwise, it is \defn{unrooted}. The structural arrangement of nodes and edges
in the tree is referred to as its \defn{topology}. The branches of the tree may
have associated lengths, representing either calendar time or genetic distance
between ancestors and their descendants.

A phylogeny with branch lengths in calendar time units is often referred to as
\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
mapped onto a timeline by using the tips of the tree, which usually map to the
present day, as a reference point~\autocite{nee1992tempo}. The corresponding
points on the timeline are called \defn{branching times}. These concepts are
illustrated in Figure~\ref{fig:speciestree}.

\begin{figure}[ht]
  \centering
  \label{fig:speciestree}
  \includegraphics{speciestree}
  \caption[Illustration of a rooted, time-scaled phylogeny]{Illustration of a
    rooted, time-scaled phylogeny. The tips of the tree, which represent extant
    taxa, are placed at the present day on the
    time axis. Internal nodes, representing extinct common ancestors to the
    extant taxa, fall in the past. The topology of the tree indicates that cats
    and dogs are the most closely related pair of species, whereas fish is most
    distantly related to any other node in the tree.}
\end{figure}

\subsubsection{Species trees and transmission trees}

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. In
general, we do not have access to the true tree relating a particular group of
extant species, as this would imply certain knowledge of speciation events in
the distant past. Therefore, all species trees considered by researchers are
estimates, based on the genetic or phenotypic similarity of the extant taxa,
the fossil record, or both. We discuss some of the challenges associated with
this estimation in subsection \ref{subsec:treeconv} below. However, for the
moment, we ignore these complications and suppose that we know the true species
tree.

The process of speciation, which causes branching in a species tree, often
occurs by an \defn{allopatric} process, where a sub-population of organisms is
isolated from the original population by a geographic
barrier~\autocite{coyne2004speciation}. Over time, the two populations diverge
genetically, eventually resulting in two distinct species.

\subsubsection{Gene trees and viral phylogenies}

\subsubsection{Using gene trees to estimate species trees}

\subsubsection{Tree shapes}

\subsection{Contact networks}

\subsubsection{Overview and epidemiology}

@@

\begin{figure}[ht]
  \centering
  \label{fig:contactnet}
  \includegraphics{contactnet}
  \caption[Illustration of a epidemic spread over a contact network]
  {Illustration of epidemic spread over a contact network. On the left, a
   contact network with five hosts, labelled A through E. Directed edges
   indicate six symmetric contacts among the hosts. Coloured, bolded edges
   represent transmissions. The epidemic began at node A, who transmitted to
   nodes B and C. Node C further transmitted to node D, and node E was not
   infected. On the right, the transmission tree topology corresponding to this
   scenario.}
\end{figure}

\subsubsection{Generative models}

%Now, consider a phylogeny not of macro-organisms, but of RNA viruses. Many of
%these, such as the human immuno-deficiency virus (HIV), evolve rapidy enough
%that populations can accumulate measurable genetic changes on time scales of
%months or years~\autocite{drummond2003measurably}. Because viruses cannot exist
%for long outside their hosts, the host replaces the outside world as the
%``environment'' in which evolution takes place. 
%
%In this light, striking parallels can be drawn between immunological and
%epidemiological processes at the host level, and the evolutionary processes
%studied in population genetics~\autocite{grenfell2004unifying}. For example,
%both the host immune response~\autocite{mcmichael2001cellular} and the presence
%of anti-retroviral drugs~\autocite{??} exert directional selection on the viral
%population within a single host. Transmission, where a sub-population of
%virions from the donor host is ``geographically'' isolated in the recpient,
%closely mirrors the allopatric speciation process and causes a branching point
%in the viral phylogeny (although this may not correspond exactly to the time of
%transmission, as discussed in the next section). Consequently, any factor which
%influences transmission dynamics, such as the virus' mode of transmission or
%the population structure of its hosts, in turn affects the shape of the
%phylogeny. The study of this interaction between host-level factors and viral
%phylogies is known as \defn{phylodynamics}~\autocite{grenfell2004unifying}. 
%
%In particular, if a time-scaled phylogeny is constructed where all viruses from
%the same host are collapsed into a single taxon, we might expect the branching
%time relating the two hosts to correspond to the time of transmission. Of
%course, this will only be the case if the two hosts' infections indeed diverged
%allopatrically at the time of transmission, and not sympatrically within the
%donor host at some previous time. This limitation, and others, in interpreting
%the branching times of the viral phylogeny as transmission times, are discussed
%in the next section. Nevertheless, even taking these limitations into account,
%the relationship between branching times in the phylogeny and transmission
%times in the population theoretically implies that the evolutionary history of
%the virus is informative about the biology and epidemiology of its hosts. In
%other words, information about, for example, transmission patterns, might be
%inferred from the viral phylogeny. Phylodynamic methods are intended to
%complement more traditional approaches to gathering epidemiological data, such
%as population surveys.
%
%
%\subsection{Sources of error in phylodynamics}
%\label{subsec:treeconv}
%
%In general, the exact topology and branch lengths of a phylogeny relating
%arbitrary taxa are not known and must be estimated. Such estimation is most
%often carried out using some type of genetic data sampled from each of the
%taxa, either amino acid sequences or gene orders. The resulting phylogeny, also
%known as a \defn{gene tree}, differs from the time-scaled \defn{species tree}
%which would contain complete information about past speciation events. In
%viral phylogenetics, the gene tree is the viral phylogeny, constructed using
%the genetic sequences of the sampled viruses. On the other hand, the species
%tree is the \defn{transmission tree}, wherein branching times correspond
%exactly to transmission times. In general, phylodynamic methods perform
%inference based on the transmission tree, which must be inferred from a viral
%phylogeny based on the available sequence data. The inference of the viral
%phylogeny itself is not an error-free process, as accurate phylogenetic
%inference is still an active area of research. Moreover, the conversion of the
%phylogeny into a transmission tree is a process which has the potential to
%introduce at least three types of error, described below. We use the terms
%``gene tree'', ``viral phylogeny'', and ``phylogeny'' interchangably, as well
%as the terms ``species tree'' and ``transmission tree''.
%
%Firstly, the branch lengths of the gene tree are measured in genetic distance,
%and must be somehow converted to units of calendar time used in the species
%tree. If we assume a constant rate of evolution along the whole tree (a
%\defn{strict molecular clock}), then this conversion is straightforward - we
%simply divide each of the branch lengths by the rate of evolution. However, it
%is well known that rates of evolution are neither constant over time nor
%constant between lineages, and the general problem of time-scaling a
%phylogenetic tree is still an active area of research. Phylodynamic inference
%relies on the assumption of a correctly time-scaled tree, and violations of
%this assumption can mislead parameter estimates.
%
%A second limitation which has perhaps greater potential to mislead phylodynamic
%inference is the potential discordance between the gene tree and the species
%tree. In an allopatric speciation, some of the genetic diversity separating the
%two eventual species may already have been present in the population
%\emph{before} the geographic isolation occured. In that case, the branching
%time in the gene tree, which indicates the time when two genetically distinct
%versions of the gene, may be earlier than the branching point in the species
%tree. In evolutionary biology, this phenomenon is referred to as
%\defn{incomplete lineage sorting}. Furthermore, if there was substantial
%variation within a population which underwent two or more speciation events in
%close succession, the order in which the species diverged may be different from
%the order in which their distinct genes arose in the original heterogeneous
%population. In this case, the topology of the species tree would be different
%from that of the gene tree. 
%
%Finally, there is the problem of rooting the tree, that is, identifying the
%node corresponding to the most recent common ancestor of all the sampled
%viruses. The choice of root influences the temporal order of the internal
%nodes in the phylogeny. If the tips are constrained to the present day, or some
%other fixed sampling time, the branch lengths in the species tree must also be
%altered to accomodate the choice of root. 
%
%To our knowledge, there has been no theoretical exploration of the quantitative
%effect of any of these sources of error on the accuracy of phylodynamic
%inference. Most phylodynamic studies assume, implicitly or explicitly, that the
%viral phylogeny provides the same information as the transmission tree, and
%errors introduced by this assumption are generally ignored. Studies employing
%Bayesian methods eliminate some of the uncertainty associated with constructing
%the viral phylogeny, but this does not account for the error introduced in the
%phylogeny-to-transmission-tree conversion process. Our work does not take a
%Bayesian approach to inferring the phylogeny, and therefore is susceptible to
%all the sources of error discussed here.
%
%\subsection{Contact networks}
%
%Compartmental models of epidemic growth and spread often assume that human
%populations are homogeneously mixed, that is, any person in the population is
%equally likely to transmit the infection to any other. More sophisticated
%models allow for the population to be divided into groups with higher or lower
%contact rates, but these still do not account for individual variation or
%population structure within compartments. In reality, and especially for
%sexually-transmitted infections such as HIV, this assumption of homogeneity is
%not realistic. There is substantial variation in the number of contacts per
%individual, and potential transmissions are limited by geographic proximity.
%The aim of using contact networks is to remove these unrealistic assumptions by 
%explicitly simulating each individual in a population, along with the contacts
%they have which could possibly lead to a transmission.
%
%A \defn{contact network} is a graphical representation of a population over
%which an epidemic can spread. Contact networks consist of nodes and edges,
%where the nodes represent people, and edges represent contacts. An edge between
%two nodes indicate that transmission is possible between the two corresponding
%individuals. For example, if we are considering an HIV transmission network,
%the edges may represent sexual contacts or individuals sharing injection drug
%paraphernalia. For an airborne pathogen such as influenza, an edge may simply
%indicate that the two people were in the same room. The edges in a contact
%network may be undirected, indicating that all contacts could possibly transmit
%to each other, or directed, indicating that transmission risk is one-way. It is
%possible to simulate an undirected network using directed edges by simply
%including two directed edges, one in either direction, for each single directed
%edge. Therefore, we exclusively consider directed contact networks here.
%
%Contact networks have been studied extensively in a variety of social,
%economic, and public health contexts. However, investigation of the effects of
%contact network structures on viral phylogenies have been much more limited.
%Existing studies have to focused on a few broad classes of contact network, and
%how phylogenies deriving from these networks have
%differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
%robinson2013dynamics}. To demonstrate our method, we focus on the same three
%contact network types as studied in~\autocite{leventhal2012inferring}, namely
%Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
%graphs. ER graphs, also known as random graphs, are the simplest generative
%graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
%which gives the probability of any edge occuring in the network. BA
%graphs~\autocite{barabasi1999emergence} model a preferential attachment process
%where nodes with high degree tend to attract more connections. They have two
%parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
%nodes with $m$ outgoing edges to the network. An existing node $v$ in the
%network is chosen to be the endpoint of these new edges with probability
%proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
%graphs~\autocite{watts1998collective}, also known as small-world graphs, have
%two parameters, $k$ and $p$. They are constructed by creating a ring lattice
%with $k$ edges per node, and then randomly rewiring each edge with probability
%$p$.
%

@@

>>>>

\subsection{Phylogenetics and phylodynamics}

@@

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa, while the \defn{internal nodes}
correspond to their common ancestors. The edges or \defn{branches} of the
phylogeny connect ancestors to their descendants. Phylogenies may have a
\defn{root}, which is a node with no descendants distinguished as the most
recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. This will be clarified below in
subsection~\ref{subsubsec:genetree}. A phylogeny with branch lengths in calendar
time units is often referred to as \defn{time-scaled}. In a time-scaled
phylogeny, the internal nodes can be mapped onto a timeline by using the tips
of the tree, which usually map to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
Figure~\ref{fig:speciestree}.

\begin{figure}[ht]
  \centering
  \label{fig:speciestree}
  \includegraphics{speciestree}
  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
    {Illustration of a rooted, ultrametric, time-scaled phylogeny. The tips of
      the tree, which represent extant taxa, are placed at the present day on
      the time axis. Internal nodes, representing extinct common ancestors to
      the extant taxa, fall in the past. The topology of the tree indicates
      that cats and dogs are the most closely related pair of species, whereas
      fish is most distantly related to any other node in the tree.}
\end{figure}

\subsubsection{Species trees and transmission trees}
\label{subsubsec:speciestree}

A \defn{species tree} is a particular type of phylogeny in which the taxa are
species, and the branching times correspond to historical speciation events. In
general, we do not have access to the true tree relating a particular group of
extant species, as this would imply certain knowledge of speciation events in
the distant past. Therefore, all species trees considered by researchers are
estimates, based on the genetic or phenotypic similarity of the extant taxa,
the fossil record, or both. We discuss some of the challenges associated with
this estimation in subsection \ref{subsubsec:treeconv} below. However, for the
moment, we ignore these complications and consider what can be said about the
true species tree.

Speciation, which causes branching in a species tree, often occurs by an
\defn{allopatric} process, where a sub-population of organisms is isolated from
the original population by a geographic barrier~\autocite{coyne2004speciation}.
Over time, differing selection pressures on either side of the barrier, coupled
with genetic drift, cause the two populations to diverge genetically. This
eventually results in two distinct species, which will appear as two subtrees
or \defn{clades} in their species tree if descendants of both survive to the
present. Of course, this is not necessarily the case, as selection may cause
one or both of the new sub-populations or their descendants to become extinct.
On the other hand, selection may favour one of the sub-populations, allowing it
to expand into new regions or niches and possibly paving the way for subsequent
speciation events. Therefore, \comment{Is this a fair statement?} we can point
out two forces primarily responsible for the shapes of species trees:
geographic barriers, which cause branching, and selection, which controls which
lineages survive to the present and influences the branching rate.

A different type of tree used in the study of pathogens is a \defn{transmission
tree}. In such trees, tips represent infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees are
generally time-scaled \comment{Is this true?} and rooted, with the root
corresponding to the index case and branching times indicating times of
transmission. The internal nodes may be labeled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the individual was
discovered or their pathogen population sampled. Consequently, the transmission
tree may not be ultrametric, but may have tips located at varying distances
from the root. Such trees are said to have \defn{heterochronous} taxa or
samples~\autocite{drummond2003measurably}, in contrast to the
\defn{isochronous} samples found in macro-organisms species trees. A
transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).

When considering epidemics of RNA viruses, transmission trees and species trees
are highly related. The viral population within a single host has often been
referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
the fast evolutionary rate of these viruses, two hosts' viral populations are
invariably genetically distinct. The transmission process is functionally
identical to allopatric speciation, involving geographic isolation of a
sub-population in a new host and subsequent genetic divergence. However, for
the most part, it is not selection but rather host epidemiology which causes
certain lineages to proliferate while driving others to extinction. A host who
recovers, dies, or is isolated equates to an extinction in the transmission
tree. On the other hand, dense populations in frequent contact often experience
rapid epidemic growth, which causes similarly rapid accumulation of branching
points in the tree.

The fact that these evolutionary (ie. speciation) and epidemiological processes
occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
is the basis for the field of
\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
transmission and speciation means that it is possible to apply many of the
tools and techniques developed for species trees to viral transmission trees,
and interpret the results in the context of transmission rates and host
epidemiology. As a basic example, the lineages-through-time
plot~\autocite{nee1992tempo}, which plots the historical speciation rate
against time, can be used to quantify the incidence of new infections over the
course of an epidemic. \comment{Reference?} In
subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
more detail, as well as several which have been developed specifically for
viral phylodynamics, and discuss what these methods can tell us about host
epidemiology.

Finally, we must note two important points. First, we have not yet considered
the selection pressures operating within a single host, such as immune
pressure, because these forces do not directly impact transmission trees unless
they cause a host to recover or die. Second, we reiterate that both
transmission trees and species trees are theoretical objects which are almost
never known with certainty for real data. Indeed, the estimation of
transmission trees is often problematic, and it is in such estimation that
within-host forces play a role. Both of these issues are discussed further
below in subsection~\ref{subsubsec:treeconv}, though we continue to ignore them
in the next subsection.

\subsubsection{Gene trees and viral phylogenies}
\label{subsubsec:genetree}

\subsubsection{Estimating transmission trees}
\label{subsubsec:treeconv}

Most applications of viral phylodynamics, including this project, aim to answer
epidemiological questions~\autocite{volz2013viral}. For example, phylodynamic
methods have been used to investigate the degree of
clustering~\autocite{hughes2009molecular} and the effect of elevated
transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
review is given below in subsection~\ref{subsubsec:appphylo}). For these
applications, we are ultimately interested in the transmission tree, which is a
representation of an epidemiological process (transmission), as opposed to the
viral phylogeny, which represents a biological process (viral evolution). As
discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees and
are in practice unknown, and must be estimated from available data. Here, we
review some of the ways to perform such estimation.

In principle, transmission trees can be reconstructed by on-the-ground
epidemiological methods, that is, by asking each person who they contacted.
This is particularly relevant for sexually transmitted diseases such as HIV,
where individuals are more likely to recall who they contacted and
when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
in detail. Most commonly, when applying phylodynamics in practice, the
transmission tree is estimated from the viral
phylogeny~\autocite{volz2013viral}. As discussed above in
subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
however, for the moment we ignore this complication and describe methods for
estimating the transmission tree if the viral phylogeny were known.

When we consider the fact that the viral phylogeny is not known with certainty,
the methods just described follow a two-step procedure. The first step is to
estimate the viral phylogeny, and the second step is to use it to estimate the
transmission tree.

\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
approach. They first estimate a time-scaled phylogeny from viral samples, then
estimate the transmissions which took place along the phylogeny. Due to
within-host evolution, the transmissions are allowed to occur anywhere along
the branches, rather than being constrained to happen at branching points.
However, the authors admit that their method requires sampling of every
infected individual, although they indicate that it could be extended to relax
this assumption.

A different approach is undertaken by Jombart
\etal~\autocite{jombart2011reconstructing}, who build transmission trees
directly from sequence data without the intermediate step of constructing a
phylogeny. Their method finds the most likely geneology among the sampled
sequences, assuming that the that the common ancestors to sampled isolates are
themselves sampled. Although this may be a realistic assumption in the early
stages of an epidemic of a slower evolving pathogen, it is unlikely for
ancestors to be present among sequences sampled from an ongoing HIV epidemic
\comment{Needs a reference}.

Several alternative approaches have been developed which aim to
incorporate available epidemiological data as well. Cottam
\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
enumerated all transmission trees consistent with this phylogeny by labelling
the internal nodes of the phylogeny in every possible way. Each of the putative
transmission trees was assigned a likelihood based on infection time and
geographic separation.

\subsubsection{Tree shapes}
\label{subsubsec:treeshape}

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often a transmission tree. Phylogenies are complex objects, and it is not
immediately obvious how to extract useful information from them with respect to
fitting a parameter. Standard statistical methods, built for numeric data
cannot be applied directly - for example, one cannot perform a regression of a
parameter of interest against a phylogeny. Therefore, before we discuss exactly
what phylodynamics can tell us about epidemics, and how it has been applied in
the past, we will review some of the methods for quantifying the shapes of
phylogenies and their similarity to each other.

\subsubsection{Applications of phylodynamics}
\label{subsubsec:appphylo}

\subsection{Contact networks}
\label{subsec:contactnet}

\subsubsection{Overview}

@@

\begin{figure}[ht]
  \centering
  \label{fig:contactnet}
  \includegraphics{contactnet}
  \caption[Illustration of a epidemic spread over a contact network]
  {Illustration of epidemic spread over a contact network. On the left, a
   contact network with five hosts, labelled A through E. Directed edges
   indicate six symmetric contacts among the hosts. Coloured, bolded edges
   represent transmissions. The epidemic began with node A, who transmitted to
   nodes B and C. Node C further transmitted to node D, and node E was not
   infected. On the right, the transmission tree topology corresponding to this
   scenario. Note that individual B was sampled before the other infected
   individuals, resulting in a tree with heterochronous tips.}
\end{figure}

\subsubsection{Generative models}
\label{subsubsec:generative}

@@

\subsection{Approximate Bayesian computation}

### 2016-01-11 doc/thesis/methods.tex
Reading and writing

\begin{figure}
  \caption{Comparison of original formulation of normalized
    lineages-through-time, developed in~\autocite{janzen2015approximate}, with
    our modified version using linear interpolation. Here, the red and blue
    trees both have uniformly spaced branching times. Using step functions
    (left), the nLTT of the two trees is non-zero due to the differing numbers
    of internal branches. Using linear interpolation, the nLTT is zero (right).
  The lines on the right graph have been offset for visibility.}
  \includegraphics[scale=0.75]{nltt}
  \label{fig:nltt}
\end{figure}

@@

\subsection{Method validation}

@@

>>>>

\begin{figure}[ht]
  \centering
  \caption[Comparison of original and modified normalised lineages-through-time]
    {Comparison of original formulation of normalized lineages-through-time,
     developed in~\autocite{janzen2015approximate}, with our modified version
     using linear interpolation. Here, the red and blue trees both have
     uniformly spaced branching times. Using step functions (left), the nLTT
     of the two trees is non-zero due to the differing numbers of internal
     branches. Using linear interpolation, the nLTT is zero (right). The lines
     on the right graph have been offset for visibility.}
  \includegraphics[scale=0.75]{nltt}
  \label{fig:nltt}
\end{figure}

@@

\subsection{Simulation experiments}

@@

Recall that our approximate Bayesian computation approach to fitting contact
network models involves simulating transmission trees under a wide variety of
parameter values, and then comparing these simulated trees to the true
transmission tree. Values which produce trees similar to the observed
transmission tree are distinguished as more likely than values which produce
trees very different from the truth. In order for this type of analysis to
succeed, it is critical that different parameter values produce different
looking trees. Otherwise, if many different values produce trees which are too
similar to each other, it will be impossible to distinguish which value is most
consistent with the real tree. Just as importantly, trees simulated with
similar parameter values must be similar to each other. In mathematical terms,
we require the trees simulated from distinct parameter values to be
\defn{separable} in tree space. The concept of separability is illustrated in
Figure~\ref{fig:separable}.

\begin{figure}[ht]
  \centering
  \includegraphics{separable}
  \caption[Separable versus non-separable pararameters in kernel space]{
    Separable versus non-separable parameters in kernel space. Trees have been
    simulated under three sets of parameters, represented as blue, red, and
    green. An observed tree is shown in black. Trees are layed out such that
    the distance between two trees corresponds to their similarity. In panel A,
    trees from the same parameter set are similar to each other, but different
    from other trees. The true tree is most consistent with the red parameters.
    In panels B and C, trees from the same parameter set are not similar to
    each other (B), or trees from different parameter sets are similar to each
    other (C). It's difficult to say which parameters the true tree is most
    consistent with.
  }
  \label{fig:separable}
\end{figure}

Before undertaking a complete ABC analysis, I analysed four simple contact
network models to determine whether their parameters could be separated in tree
kernel space. The four models are described in detail in
subsection~\ref{subsubsec:generative} of the introduction. Briefly, they are:
random networks, where each possible contact has a fixed probability of
occuring; preferential attachment networks, where highly connected nodes tend
to attract more contacts; small world networks, where nodes are connected to
their immediate neighbours and the occasional far-flung contact; and full
networks, where every possible connection is present. I will describe here only
the procedure and results for the preferential attachment networks. The details
of the other three types of graph can be found in the supplemental materials.

The method of testing for separability was described previously
in~\autocite{poon2015phylodynamic}, but I will reiterate it here for
completeness. As a concrete example, consider the attachment power parameter
$\alpha$ of the preferential attachment networks. This parameter describes the
strength of attraction to highly connected nodes and is bounded below by zero
indicating no extra attraction. By qualitative observation, I determined that a
power of 2.0, which produced networks with very few ``hub'' nodes with
extremely high degree, was a suitable upper bound (see
Figure~\ref{fig:pabounds}). Therefore, I chose to test the values 0.5, 1.0, and
1.5 for separability. The other parameters were fixed: the number of nodes in
the network was 5000, and the mean degree of each node in the network was four.
As discussed in subsection~\ref{subsubsec:generative}, this is the smallest
mean degree value for preferential attachment networks which produces networks
which are more than trees.

\begin{figure}[ht]
  \centering
  \label{fig:pabounds}
  \includegraphics{pa_power_bounds}
  \caption[Upper and lower bounds on preferential attachment power]{
    Qualitative justification for choice of zero and two as lower and upper
    bounds on preferential attachment power $\alpha$. (A) Preferential
    attachment network on 50 nodes with $\alpha = 0$, the lower bound enforced
    by the model. (B) Preferential attachment network on 50 nodes with $\alpha
    = 2$, where a few nodes have very high degree but the majority have very
    low degree. (C) Density plot of node degrees in a 5000-node network with
    $\alpha = 0$. The maximum degree of a node in this network was 24. (D)
    Density plot of node degrees in a 5000-node network with $\alpha = 2$. The
    maximum degree of a node in this network was 4942. 
  }
\end{figure}

For each of the values of $\alpha$, I generated 100 networks on 5000 nodes. An
epidemic was simulated over each network (see subsection X) until 1000 nodes
were infected, and 500 of those infected nodes were sampled to form a
transmission tree. This resulted in 300 total simulated transmission trees -
100 for each of the three values of $\alpha$. Next, I computed the tree
kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
matrix, where the value at the $(i, j)$th position was the tree kernel of the 
$i$th and $j$th trees.

The tree kernel provides a pairwise similarity score between two trees. The
higher the kernel score, the more similar the trees are to each other.
Therefore, to have separability, we need trees simulated with the same value of
$\alpha$ to have high kernel scores with each other, but low kernel scores with
trees from different $\alpha$ values. We can visually check whether or not this
is true by laying out the trees as points on a graph, in such a way that trees
with high scores are close to each other, but trees with low scores are far
apart. This is accomplished by performing a kernel principal components
analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
representation of points in a high dimensional space which preserves as much of
the variation in the data as possible. kPCA performs the same task, but using
the dot products of each pair of points as input, instead of the data points
themselves. A two-dimensional kPCA projection of the simulated trees is shown
in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
infected nodes - is the central panel. To ensure that the method could be used
in a variety of contexts, the same analysis was performed with 500 and 2000
infected nodes, as well as 100 and 1000 sampled tips. The projections show that
all three values are well separated from each other in many of the scenarios.
With smaller trees, it becomes harder to disinguish $\alpha = 0.5$ from $\alpha
= 1.0$ in two dimensions.

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:pacrossv}
  \includegraphics{kernel_pa_crossv}
  \caption[Cross-validation performance of kernel support vector machine
  classifier for preferential attachment power]{
    Cross-validation performance of kernel support vector machine classifier
    for preferential attachment power.
  }
\end{figure}
